 Allocation scoring rules as a generalization of quantile scoring rules

### 2023-06-14 ISF_talk/talk.qmd
more entropy/divergence work.  getting slide started.

>>>>

## The backstory

Over last decade, multicenter effort to collect forecasts via hubs for ID outbreaks (epi/pan/endemic)
with central purpose to "inform public health responses"

With uncertainty quantification being well-established as needed DM input, 
hubs have trended toward eliciting probabilistic forecasts

Basic element of any such response is an *allocation decision* regarding
 - medical supplies
 - personnel 
 - facility capacity
 - R&D funding
 - public tolerance for NPIs

Allocations can be wrt space, time, demographic

Similar to other contexts where there are large-scale institutional efforts to use expert judgment to assess and manage social risk
 - finance
 - meteorology
 - military conflict

Even with complex game-theoretic approachs to collective decision making,
*allocation* almost always a requirement to strategy development

## Scoring rules and social utility

Hubs have tried to use SRs to rank and combine forecasts in ways that offer best suppport for PH responders

This can be cast in decision-theoretic framework of utility, loss functions, and Bayes acts.

The DM needs to decide what action $a$ to take regarding an uncertain outcome $Y$ such as the number of severe cases in various locations.
Suppose the DM also has
a loss function $l(a,y)$, e.g., the unmet need when $a$ beds are allocated for $y$ severe cases,
and a utility function U quantifying, e.g., the ratio of the value of preventing 1000 unhospitalized severe cases versus the value of preventing 10 unhospitalized severe cases.  That is U specifies numbers such as U(l(10,20)) / U(l(1000,2000)).
Then the Bayes act relative to a forecast $Y \sim F$ is the action $a^F$ that will maximize the DM’s expected utility assuming that in reality $Y \sim F$.

## Basic computational approach

Adopt an iterative Lagrange multiplier method following long history in inventory management of solving
*constrained multi-product newsvendor* problems

### 2023-06-14 notes/allocation_notes.rmd
more entropy/divergence work.  getting slide started.

Let $F,F_i$ be general distributions such as forecaster beliefs or reports about $Y$, and $G$ be the true distribution of $Y$. 
For a general scoring rule $S$, we define the $S$-entropy of $F$ as 
\begin{align}
e_S(F) &:= S(F, F) = E_F[S(F, Y)]
\end{align}
and the $S$-entropy of the random variable $Y$ as
\begin{align}
e_S(Y) &:= e_S(G).
\end{align}
Applying this to $S_\alpha$, we can define the $g,\alpha$-entropy of $F$ as
\begin{align}
e_{\alpha}(F) &:= S_{\alpha}(F, F) = E_F[s_{\alpha}(q_{F,\alpha}, Y)] .
\end{align}
A forecaster believing that $Y \sim F$ will expect to receive a score of $e_{\alpha}(F)$ by reporting their $\alpha$ quantile, while an
omniscient forecaster \emph{knowing} that $Y\sim G$ will expect a score of $e_S(Y)$ for reporting $q_{\alpha,Y}$.
According to \eqref{eqn:scalar_obj_cdf_form} and \eqref{eqn:obj_pexp},
\begin{align}
e_{\alpha}(F)
&= (1-\alpha) \int_{-\infty}^{q_{\alpha,F}} F(u) dg(u) + \alpha \int_{{q_{\alpha,F}}}^{\infty}(1-F(u)) dg(u) \\
&= \alpha \overline{g}\sb{F}-\overline{g}_{F}(q_{\alpha,F}).
\end{align}
Similarly, we define the $g,\alpha$-divergence between two distributions as
\begin{align}
d_{\alpha}(F_1,F_2) &:= S_{\alpha}(F_1,F_2) - S_{\alpha}(F_2,F_2) = S_{\alpha}(F_1,F_2) - e_{\alpha}(F_2). \label{egn:divergence}
\end{align}
Again from \eqref{eqn:scalar_obj_cdf_form} we get
\begin{align}
d_{\alpha}(F_1,F_2)
&= (1-\alpha) \int_{-\infty}^{q_{\alpha,F_1}} F_2(u) dg(u) + \alpha \int_{{q_{\alpha,F_1}}}^{\infty}(1-F_2(u)) dg(u) \\
&-(1-\alpha) \int_{-\infty}^{q_{\alpha,F_2}} F_2(u) dg(u) - \alpha \int_{{q_{\alpha,F_2}}}^{\infty}(1-F_2(u)) dg(u) \\
&= (1-\alpha) \int_{q_{\alpha,F_2}}^{q_{\alpha,F_1}} F_2(u) dg(u) + \alpha \int_{{q_{\alpha,F_1}}}^{q_{\alpha,F_2}}(1-F_2(u)) dg(u).
\end{align}
And from \eqref{eqn:obj_pexp},
\begin{align}
d_{\alpha}(F_1,F_2) &= g(q_{\alpha,F_1})(F_2(q_{\alpha,F_1}) - \alpha) -\overline{g}_{F_2}(q_{\alpha,F_1}) + \alpha \overline{g}\sb{F_2}  - \alpha \overline{g}\sb{F_2} + \overline{g}_{F_2}(q_{\alpha,F_2}) \\
&= g(q_{\alpha,F_1})(F_2(q_{\alpha,F_1}) - \alpha) + \int_{q_{\alpha,F_1}}^{q_{\alpha,F_2}}g(y)dF_2(y)
\end{align}
With \eqref{egn:divergence}, we get 
\begin{align}
S_{\alpha}(F,G) = d_{\alpha}(F,G) + e_{\alpha}(Y), \label{eqn:marg_score_decomp}
\end{align}
expressing the true expected score of $F$ 
(i.e., not that calculated by the forecaster) as the sum of the failure of $F$ to match $G$ and the inherent uncertainty of $Y$.

@@

<!-- {}
\begin{align}{}
\overline{s}_{F,\kappa,\alpha}(x) &:= s_{\kappa,\alpha}(x,F) := E_F[s_{\kappa,\alpha}(x,Y)] \\
\overline{s}_{\kappa,\alpha}(x) &:= s_{\kappa,\alpha}(x,G) \\
s_{\symbfit{\kappa},\symbfit{\alpha}}(\mathbf{x}, \mathbf{F}) &:= 
\overline{s}_{\mathbf{F},\symbfit{\kappa},\symbfit{\alpha}}(\mathbf{x}) = 
\sum_{i}^{\mathrm{dim}\mathbf{x}} s_{\kappa_i,\alpha_i}(x_i, F_i) \\
S(F,y) &:= s(q_{F,\alpha}, y) \\
S(F_1,F_2) &:= E_{F_2}[S(F_1,Y)] = s(q_{F_1,\alpha}, F_2) \\
\overline{S}(F) &:= E_G[s(q_{F, \alpha}, Y)]
\end{align} 
-->

@@

In the inventory management literature, minimization of an expectation of the form of $\sbar_F(x)$ for a given $F$ arises in the \emph{newsvendor problem}.  This involves the ordering decision faced by a retailer of a perishable good (such as newpapers) when customer demand is uncertain. Because the focus here is on revenues and expenditures, terms other than just the over- and underprediction penalties in \eqref{eqn:scalar_score} often appear, giving a retailer's scoring function as
\begin{align}
s_{\mathbf{n}}(x,y) & = c x - p \min(x,y) - u(x-y)\sb{+} + r(y-x)\sb{+} \label{eqn:news_score} \\
&= \begin{cases}
cx - px + r(y-x), & x \leq y \\
cx - py - u(x-y), & x > y
\end{cases}
\end{align}
where $\mathbf{n} = \{c,p,u,r\}$.
This score is netting the monetary transfers that occur when
where $x$ newspapers are ordered at price $c$, $\min(y,x)$ are sold at retail price $p$, $(x-y)\sb{+}$ are sold at salvage price $u$, 
and the "goodwill", monetarily valued at $r$, of $(y-x)\sb{+}$ customers is lost. 
Using $x = y + (x-y)_+ - (y-x)_+$ and $\min(x,y) = y - (y-x)_+$ \eqref{eqn:news_score} can be rewritten as
\begin{align}
s_{\mathbf{n}}(x,y) &= (c-u)(x-y)\sb{+} + (p-c+r)(y-x)\sb{+} -(p-c)y
\end{align}
giving the retailer's objective function under a demand distribution $F$ as
\begin{align}
\sbar_{F, {\mathbf{n}}}(x) = (c-u)E_F[(x-Y)\sb{+}] + (p-c+r)E_F[(Y-x)\sb{+}] -(p-c)E_F[Y].
\end{align}
This differs by a constant from our $\sbar_{F,O,U}(x)$ in \eqref{eqn:scalar_obj} with $g(x)=x$, $O=c-u$ 
and $U=p-c+r$, and so has the same minimizer   
\begin{align}
q_{F,O,U} = F^{-1}\left(\frac{U}{O+U}\right) = F^{-1}\left(\frac{p-c+r}{p-u + r}\right).
\end{align} 
Note, for reference, that in the literature $\sbar_{F, {\mathbf{n}}}(x)$ is often encountered in the form
\begin{align}
&(O+U) E_F[(x-Y)\sb{+}] + U E_F[Y] E_F[(Y-x)\sb{+}-(x-Y)\sb{+}] -(p-c)E_F[Y] \\
&(O+U) E_F[(x-Y)\sb{+}] + U (E_F[Y] - x) -(p-c)E_F[Y]\\
=\,&(p+r-u) \int_{0}^{x} (x - y)dF(y) - (p-c+r)x+ rE_F[Y]
\end{align}

>>>>

Let $F,F_i$ be general distributions such as forecaster beliefs or reports about $Y$, and $G$ be the true distribution of $Y$. 
For a general scoring rule $S$, we define the $S$-entropy of $F$ as 
\begin{align}
e_S(F) &:= S(F, F) = E_F[S(F, Y)]
\end{align}
and the $S$-entropy of the random variable $Y$ as
\begin{align}
e_S(Y) &:= e_S(G).
\end{align}
Applying this to $S_\alpha$, we can define the $g,\alpha$-entropy of $F$ as
\begin{align}
e_{\alpha}(F) &:= S_{\alpha}(F, F) = E_F[s_{\alpha}(q_{\alpha,F}, Y)] .
\end{align}
A forecaster believing that $Y \sim F$ will expect to receivq_e a score of $e_{\alpha}(F)$ by reporting their $\alpha$ quantile, while an
omniscient forecaster \emph{knowing} that $Y\sim G$ will expect a score of $e_S(Y)$ for reporting $q_{\alpha,Y}$. Note that if 
$F_{y_0} = \mathbb{1}\{x \geq y_0\}$ is deterministic, then 
\begin{align}
e_{\alpha}(F_{y_0}) = E_{F_{y_0}}[s_{\alpha}(q_{\alpha, F_{y_0}}, Y)] = s_{\alpha}(y_0, y_0) = 0. \label{eqn:entropy_determ}
\end{align}
Similarly, if $F_{\mathbf{y}_n} = \mathrm{Unif}\{y_1, \ldots, y_n\}$ is the empirical distribution of a sample $\mathbf{y}_n = \{y_1, \ldots, y_n\}$, then
\begin{align}
e_{\alpha}(F_{\mathbf{y}_n}) = E_{F_{\mathbf{y}_n}}[s_{\alpha}(q_{\alpha, F_{\mathbf{y}_n}}, Y)] 
& = \frac{1}{n}\sum_{i=1}^{n} s_{\alpha}(y_{\lceil \alpha n \rceil}, y_i) \\
& = \frac{1}{n} \left\{(1-\alpha)\sum_{i=\lceil \alpha n \rceil}^{n}g(y_i) - g(y_{\lceil \alpha n \rceil}) + 
\alpha \sum_{i=1}^{\lceil \alpha n \rceil} g(y_{\lceil \alpha n \rceil}) - g(y_i) \right\}
\end{align}
which we also call the $g,\alpha$-entropy $e_{\alpha}(\mathbf{y}_n)$ of the sample $\mathbf{y}_n$.

According to \eqref{eqn:scalar_obj_cdf_form} and \eqref{eqn:obj_pexp},
\begin{align}
e_{\alpha}(F)
&= (1-\alpha) \int_{-\infty}^{q_{\alpha,F}} F(u) dg(u) + \alpha \int_{{q_{\alpha,F}}}^{\infty}(1-F(u)) dg(u) \\
&= \alpha \overline{g}\sb{F}-\overline{g}_{F}(q_{\alpha,F}).
\end{align}

Similarly, we define the $g,\alpha$-divergence between two distributions as
\begin{align}
d_{\alpha}(F_1,F_2) &:= S_{\alpha}(F_1,F_2) - S_{\alpha}(F_2,F_2) = S_{\alpha}(F_1,F_2) - e_{\alpha}(F_2). \label{egn:divergence}
\end{align}
If $F_2 = F_{y_0}$ is deterministic, then from \eqref{eqn:entropy_determ} we have 
\begin{align}
d_{\alpha}(F_1, F_{y_0}) = S_{\alpha}(F_1, F_{y_0}) = E_{F_{y_0}}[S_{\alpha}(F_1, Y)] = S_{\alpha}(F_1, y_0) = s_{\alpha}(q_{\alpha, F_1}, y_0)
\end{align}
whereas if $F_1 = F_{y_0}$, we get
\begin{align}
d_{\alpha}(F_{y_0}, F_2) = E_{F_2}[s_{\alpha}(y_0,Y)] - e_{\alpha}(F_2).
\end{align}
When $F_2 = F_{\mathbf{y}_n}$ is an empirical distribution, then 
\begin{align}
d_{\alpha}(F_1, F_{\mathbf{y}_n}) 
&= E_{F_{\mathbf{y}_n}}[S_{\alpha}(F_1, Y)] - e_{\alpha}(\mathbf{y}_n) \\
&= \frac{1}{n}\sum_{i=1}^{n} s_{\alpha}(q_{\alpha,F_1}, y_i) - s_{\alpha}(y_{\lceil \alpha n \rceil}, y_i)
\end{align}
whereas inserting the $F_{\mathbf{y}_n}$ in the first position gives
\begin{align}
d_{\alpha}(F_{\mathbf{y}_n}, F_2) = E_{F_2}[s_{\alpha}(y_{\lceil \alpha n \rceil},Y)] - e_{\alpha}(F_2).
\end{align}

<!--
&= (1-\alpha)(g(q_{\alpha, F_1}) - g(y_{\lceil \alpha n \rceil})) \\
&+ \frac{1}{n} \left\{\sum_{i \in I_1} g(y_i) - g(q_{\alpha, F_1})
+\sum_{i \in I_2} g(q_{\alpha, F_1}) - g(y_i) 
+\sum_{i \in I_3} g(y_{\lceil \alpha n \rceil})) - g(q_{\alpha, F_1}) \right\}
 \text{ using \eqref{eqn:scalar_score_OR}}
where $I_1 = \{i | q_{\alpha, F_1} \leq y_i \leq y_{\lceil \alpha n \rceil}\}$, 
$I_2 = \{i | y_{\lceil \alpha n \rceil} \leq y_i \leq q_{\alpha, F_1}\}$, and 
$I_3 = \{i | y_{\lceil \alpha n \rceil}, q_{\alpha, F_1} \leq y_i\}$.
-->

As for entropy, we also can get more explicit forms for the divergence in terms of distributions $F_1, F_2$. 
From \eqref{eqn:scalar_obj_cdf_form} we get
\begin{align}
d_{\alpha}(F_1,F_2)
&= (1-\alpha) \int_{-\infty}^{q_{\alpha,F_1}} F_2(u) dg(u) + \alpha \int_{{q_{\alpha,F_1}}}^{\infty}(1-F_2(u)) dg(u) \\
&-(1-\alpha) \int_{-\infty}^{q_{\alpha,F_2}} F_2(u) dg(u) - \alpha \int_{{q_{\alpha,F_2}}}^{\infty}(1-F_2(u)) dg(u) \\
&= (1-\alpha) \int_{q_{\alpha,F_2}}^{q_{\alpha,F_1}} F_2(u) dg(u) + \alpha \int_{{q_{\alpha,F_1}}}^{q_{\alpha,F_2}}(1-F_2(u)) dg(u).
\end{align}
And from \eqref{eqn:obj_pexp},
\begin{align}
d_{\alpha}(F_1,F_2) &= g(q_{\alpha,F_1})(F_2(q_{\alpha,F_1}) - \alpha) -\overline{g}_{F_2}(q_{\alpha,F_1}) + \alpha \overline{g}\sb{F_2}  - \alpha \overline{g}\sb{F_2} + \overline{g}_{F_2}(q_{\alpha,F_2}) \\
&= g(q_{\alpha,F_1})(F_2(q_{\alpha,F_1}) - \alpha) + \int_{q_{\alpha,F_1}}^{q_{\alpha,F_2}}g(y)dF_2(y).
\end{align}
This offers a more explicit form of the divergence of a distribution from s sample $\mathbf{y}_n$:
\begin{align}
d_{\alpha}(F_1,F_{\mathbf{y}_n}) 
= g(q_{\alpha,F_1})\left(\frac{1}{n}\sum_{1=1}^n \mathbb{1}\{y_i \leq q_{\alpha,F_1}\} - \alpha\right) 
+ \frac{1}{n}\sum_{1=1}^{\lceil \alpha n \rceil} \mathbb{1}\{y_i > q_{\alpha,F_1}\}g(y_i).
\end{align}

With \eqref{egn:divergence}, we get 
\begin{align}
S_{\alpha}(F,G) = d_{\alpha}(F,G) + e_{\alpha}(Y), \label{eqn:marg_score_decomp}
\end{align}
expressing the true expected score of $F$ 
(i.e., not that calculated by the forecaster) as the sum of the failure of $F$ to match $G$ and the inherent uncertainty of $Y$.

@@

<!-- {}
\begin{align}{}
\overline{s}_{F,\kappa,\alpha}(x) &:= s_{\kappa,\alpha}(x,F) := E_F[s_{\kappa,\alpha}(x,Y)] \\
\overline{s}_{\kappa,\alpha}(x) &:= s_{\kappa,\alpha}(x,G) \\
s_{\symbfit{\kappa},\symbfit{\alpha}}(\mathbf{x}, \mathbf{F}) &:= 
\overline{s}_{\mathbf{F},\symbfit{\kappa},\symbfit{\alpha}}(\mathbf{x}) = 
\sum_{i}^{\mathrm{dim}\mathbf{x}} s_{\kappa_i,\alpha_i}(x_i, F_i) \\
S(F,y) &:= s(q_{\alpha, F}, y) \\
S(F_1,F_2) &:= E_{F_2}[S(F_1,Y)] = s(q_{\alpha, F_1}, F_2) \\
\overline{S}(F) &:= E_G[s(q_{ \alpha,F}, Y)]
\end{align} 
-->

@@

In the inventory management literature, minimization of an expectation of the form of $\sbar_F(x)$ for a given $F$ arises in the \emph{newsvendor problem}.  This involves the ordering decision faced by a retailer of a perishable good (such as newpapers) when customer demand is uncertain. Because the focus here is on revenues and expenditures, terms other than just the over- and underprediction penalties in \eqref{eqn:scalar_score} often appear, giving a retailer's scoring function as
\begin{align}
s_{\mathbf{n}}(x,y) & = c x - p \min(x,y) - u(x-y)\sb{+} + r(y-x)\sb{+} \label{eqn:news_score} \\
&= \begin{cases}
cx - px + r(y-x), & x \leq y \\
cx - py - u(x-y), & x > y
\end{cases}
\end{align}
where $\mathbf{n} = \{c,p,u,r\}$.
This score is netting the monetary transfers that occur when
where $x$ newspapers are ordered at price $c$, $\min(y,x)$ are sold at retail price $p$, $(x-y)\sb{+}$ are sold at salvage price $u$, 
and the "goodwill", monetarily valued at $r$, of $(y-x)\sb{+}$ customers is lost. 
Using $x = y + (x-y)_+ - (y-x)_+$ and $\min(x,y) = y - (y-x)_+$ \eqref{eqn:news_score} can be rewritten as
\begin{align}
s_{\mathbf{n}}(x,y) &= (c-u)(x-y)\sb{+} + (p-c+r)(y-x)\sb{+} -(p-c)y
\end{align}
giving the retailer's objective function under a demand distribution $F$ as
\begin{align}
\sbar_{F, {\mathbf{n}}}(x) = (c-u)E_F[(x-Y)\sb{+}] + (p-c+r)E_F[(Y-x)\sb{+}] -(p-c)E_F[Y].
\end{align}
This differs by a constant from our $\sbar_{F,O,U}(x)$ in \eqref{eqn:scalar_obj} with $g(x)=x$, $O=c-u$ 
and $U=p-c+r$, and so has the same minimizer   
\begin{align}
q_{O,U, F} = F^{-1}\left(\frac{U}{O+U}\right) = F^{-1}\left(\frac{p-c+r}{p-u + r}\right).
\end{align} 
Note, for reference, that in the literature $\sbar_{F, {\mathbf{n}}}(x)$ is often encountered in the form
\begin{align}
&(O+U) E_F[(x-Y)\sb{+}] + U E_F[Y] E_F[(Y-x)\sb{+}-(x-Y)\sb{+}] -(p-c)E_F[Y] \\
&(O+U) E_F[(x-Y)\sb{+}] + U (E_F[Y] - x) -(p-c)E_F[Y]\\
=\,&(p+r-u) \int_{0}^{x} (x - y)dF(y) - (p-c+r)x+ rE_F[Y]
\end{align}
